---
title: "Index Standardization Overview"
author: "Grant Adams"
format: pptx
editor: visual
---

# Index standardization

Many fisheries rely on development of indices that track abundance or biomass under the assumption:

$$
I_y = q *N_y
$$

$I_y$ = annual index

$q$ = catchability (fraction of population captured per unit effort)

$N_y$ = Numbers or density

------------------------------------------------------------------------

# Sources of data

Scientific surveys

-   Surveys designed to limit variation in $q$
-   Developed for highly valuable or important species (e.g. anchovy, pollock, whales)

Fishery-dependent data

-   $q$ is dependent on fishing behavior and fleet dynamics
-   Surveys are not practical or expensive (e.g. tuna fisheries)

------------------------------------------------------------------------

# Common index models

-   Generalized Linear Models (GLMs)

-   Generalized Linear Mixed Models (GLMMs)

-   Generalized Additive Models (GAMs)

-   Generalized Additive Mixed Models (GAMMs)

Model choice is often dependent on data-available and analyst

------------------------------------------------------------------------

# Fundamentals

$$
\mu_i=E(I_{i})
$$

$$
g(\mu_i)= \text{X}_i * \beta
$$

$\mu_i$ = expected value

$I_i$ = observed value $i$

$g()$ = link function (log, logit, etc)

$X_i$ = Explanatory variables (year, month, vessel, etc)

$\beta$ = estimated parameters

------------------------------------------------------------------------

# Common distributions

*Presence/absence*:

-   Binomial (logit link function)

*Numbers*:

-   Poisson (log link function)

-   Negative binomial (log link function)

*Biomass*:

-   Normal

-   Gamma

Many more!!

------------------------------------------------------------------------

# Common covariates $X$

-   Year or time

Factors that influence catchability:

-   Vessel (some vessels may be more efficient/bigger nets)

-   Time of day (fish move up and down the water column)

-   Temperature (fish migrate into the survey area seasonally with temperature)

------------------------------------------------------------------------

# Implementation (1)

Multiple R packages

GLM: `stats`

GLMM: `glmmTMB`, `lme4`, `nlme`

GAM: `mgcv`

GAMM: `mgcv`, `gamm4`

------------------------------------------------------------------------

# GLM implementation (2)

```{r, echo = TRUE, eval = FALSE}
mod = glm(CPUE ~ Year + Depth, data = mydata, family = poisson(link = "log"))
```

Distribution = `family`

Link function $f()$ = `link`

Covariates $X$ = `Year + Depth`

------------------------------------------------------------------------

# Diagnostics

Standard regression diagnostics should be used such `plot(mod)`

-   Residuals vs fitted values (model misspecification)

-   Sqrt of abs of residuals vs expected values (variance correctly specified)

-   QQplot (residuals normally distributed)

-   Observed vs predicted

------------------------------------------------------------------------

# Good model diagnostics

```{r}
library(mgcv)
set.seed(6)
dat <- gamSim(1,n=2000,dist="poisson",scale=.1)

## use "cr" basis to save time, with 2000 data...
b2<-gam(y~s(x0,bs="cr")+s(x1,bs="cr")+s(x2,bs="cr")+
            s(x3,bs="cr"),family=poisson,data=dat,method="REML")
par(mfrow = c(2, 2), mar = c(3, 2, 1, 1))
gam.check(b2)
```

------------------------------------------------------------------------

# Wrong distribution

```{r}
## use "cr" basis to save time, with 2000 data...
b3<-gam(y~s(x0,bs="cr")+s(x1,bs="cr")+s(x2,bs="cr")+
            s(x3,bs="cr"),data=dat,method="REML")
par(mfrow = c(2, 2), mar = c(3, 2, 1, 1))
gam.check(b3)
```

# Missing predictors

```{r}
## use "cr" basis to save time, with 2000 data...
b4<-gam(y~s(x0,bs="cr"),family = poisson, data=dat,method="REML")
par(mfrow = c(2, 2), mar = c(3, 2, 1, 1))
gam.check(b4)
```
